Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging

### 2013-03-06 HRUIonGPU/jrnl_article.tex

%\documentclass[journal]{IEEEtran}

% Single column review mode:
\documentclass[12pt,journal,onecolumn]{IEEEtran}
\newcommand\CLASSINPUTbaselinestretch{2}
\RequirePackage{calc}
\RequirePackage{fp} 
% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
% Adding the 2cm gutter width to the odd/even side margins:
\setlength\hoffset{0pt}
\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side

@@

\begin{abstract}
%\boldmath
Capon beamforming is associated with a high computational complexity, which limits its use as a real-time array processor in many applications. In this paper we present an implement of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting. To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the number of array elements increases. For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging is shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters. Image quality is investigated in videos of both a simulated phantom and an \textit{in-vivo} cardiac dataset. These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

% denne må utelates i IUS-artikkelen
In recent work \cite{asen2012}(cite JOE or ECUA?) we have applied a GPU-based Capon beamformer for both active sonar and for cardiac ultrasound imaging. Sonar has a real-time requirement less strict than medical ultrasound imaging, which means that most array configurations can be handled in real-time using the GPU. Medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates at a level where even a high-end GPU can not do the processing required for a straightforward implementation of the Capon beamformer. Since the execution time is a cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming have therefore recently been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we will make use of a beamspace transformation described by Nilsen et al. \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases. The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

@@

\begin{figure}
\centerline{\includegraphics[width=3in]{gfx/gpu_layout_vertical_lr.png}}
\caption{Calculation of Capon adaptive weights mapped to GPU architecture. A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups. Note that several samples can be processed per block of threads if $L$ is small. The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
\label{fig:gpulayout}
\end{figure}

@@

To reach this goal we start by showing how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
In time we have that
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculate from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ (cite JOE?). However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$ (if $L=M/2$). Note that sliding will break independence along the sliding dimension, thus, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is increased to the maximum number of threads per block (currently 1024) with this approach. However, for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.

Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated. However, the GPU computes $\mat{\hat{R}}$ in such a way that the lower triangle can be found with little overhead (cite JOE). This also means that no special solver is needed in order to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if a symmetric solver could have saved some instructions.

@@

\begin{figure}
%\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
%\hfil
%\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_606.png}%
%\label{fig:benchUS64}}}
\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_606.png}}
\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15$ cm deep sector. Execution time is plotted as a function of subarray length $L$. The plot shows execution times for the three steps listed in Section \ref{sec:meth}. Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
\label{fig:bench}
\end{figure}

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=606_Nb=3.png} \label{fig:benchBS64}}
\hfil
\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=606_Nb=3.png}%
\label{fig:benchBS96}}}
\caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$ cm deep sector. Execution time is plotted as a function of subarray length $L$.  If $L$ had been kept constant on $M/2$ and execution time was plotted against $N_b$, the plot would be similar to the one in Fig. \ref{fig:bench}. The variation in execution time is caused by adaptive kernel launch configurations.}
\label{fig:benchBS}
\end{figure*}
In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth). Problem sizes are calculated based on 32, 64, and 96 element phased arrays with a 2.5 MHz center frequency. Assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz, and when imaging down to a depth of 15 cm we have approximately 400 samples per range line. To correctly sample a $70\degree$ sector with 32, 64 and 96 element arrays ($\lambda/2$ pitch), we need at least 40, 80 and 120 beams respectively. The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in GPU memory. What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.

A benchmark of the implementation described in Section \ref{sec:meth} is presented in Fig. \ref{fig:bench}. Benchmarks are based on 32 and 64 elements phased arrays respectively, using 40 and 80 transmit beams. For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$. For $M=64$, $L=32$ and $K=2$ (Fig. \ref{fig:benchUS64}), a typical setup for high quality cardiac imaging, processing throughput is 10 frames per second.

In Fig. \ref{fig:benchBS} we present similar benchmarks for the beamspace version of the Capon beamformer introduced in Section \ref{sec:bs}. Benchmarks are based on 64 and 96 elements phased arrays, using 80 and 120 transmit beams. The figure only present results with use of temporal averaging ($K=2$). For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 100 frames per second. For all four benchmarks there is a line indicating the time it takes to acquire the underlying image using single line acquisition. Hence, below this line we are capable of performing real-time processing when a single receive line is used per transmit line.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=3in]{gfx/simulation_slice.png} \label{fig:simulationSlice}}
\hfil
\subfloat[]{\includegraphics[width=3in]{gfx/invivo_slice.png} \label{fig:invivoSlice}}
}
\caption{Lateral slices taken from the images in Fig. \ref{fig:phantom} and Fig. \ref{fig:invivo}. a) From Fig. \ref{fig:phantom} at 93 mm range. Straight through the two bright point scatterers. b) From Fig. \ref{fig:invivo} at 96 mm range, going through the tip of the mitral valve.}
\label{fig:slices}
\end{figure*}

To investigate the image quality of both the Capon and BS Capon beamformer a phantom was simulated using Field II \cite{Jensen1992} (Fig. \ref{fig:phantom}). The simulation was performed using a $\lambda/2$ pitch, 10 $\mu$m kerf, 50\% bandwidth and 10 mm high phased array. The transducer was excited with a 2.5 MHz and $3\lambda$ long pulse with transmit focus at 80 mm. The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB. Speckle level is at 0 dB. The resulting image has been processed both using DAS with uniform weights, Capon and BS Capon. The presented image is the first frame in a series of images where motion is imposed on both circles, point targets and the probe. The entire simulation is presented in an attached video.

@@

To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Fig. \ref{fig:phantom} and \ref{fig:invivo} are included in Fig. \ref{fig:slices}. For the phantom, the cut goes through the two point targets at 93 mm depth, and for the cardiac image it goes through the tip of the mitral valve at 96 mm depth.

@@

\section{Trading resolution for speed}\label{sec:trade}
As seen from Fig. \ref{fig:bench} and \ref{fig:benchBS}, execution time is reduced when either $L$ or $N_b$ is reduced. So to gain an  increase in processing throughput we can reduce $L$ or $N_b$. The question is then, how does this impact image quality? In Fig. \ref{fig:speed_res_trade}, the minimum distance between the two point targets in Fig. \ref{fig:phantom} (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with Capon) and $N_b$ (when processed with BS Capon). For the result in Fig. \ref{fig:speed_res_trade}, $d=0.01$ has been used on both methods.

For high $L$ and $N_b$ values the two methods are equal. However, for lower values, BS Capon provides higher resolution than what is possible with small subarrays. We see that the resolution for BS Capon gradually decreases due to an automatically increase in diagonal loading (See end of Section \ref{sec:bs}). If this is compensated for, BS Capon will provide resolution almost equal to Capon for all values except $N_b=1$ \cite{Nilsen2009}. For $N_b=1$ the resolution is worse than for DAS because of a triangular apodization caused by subarray averaging. In element space, resolution decays rapidly as a result of a reduced effective adaptive aperture. For $N_b=3$, BS Capon still provides approximately 1 mm improved resolution over DAS. However, note that resolution for the Capon beamformer highly depends on image SNR and the choice of parameters. Thus, different SNR scenarios and choices of parameters will give different results. It is also worth noting that Capon beamforming in this example performs worse than DAS for small subarrays. For small values of $L$, there are not enough degrees of freedom in order steer zeros at each point target.

@@

\section{Discussion}\label{sec:dis}
In this paper we have presented a parallel implementation of the Capon beamformer in both element- and beam-space for high resolution cardiac ultrasound imaging. The results show that it is now possible to apply BS Capon in real-time in cardiac ultrasound imaging by utilizing the power of GPUs. According to Fig. \ref{fig:benchBS} it is also possible to process two parallel receive lines per transmit (for both $M=64$ and $M=96$. $L=M/2$). However, these extra lines might be needed for increased sampling instead of increased frame rate when using the Capon beamformers. If parallel receive lines are used to obtain denser sampling, acquisition time remains the same but processing times are increased with a factor equal to the amount of oversampling. If parallel receive lines are used for increased frame rate, the acquisition time is lowered, and the processing time remains constant. Hence, four parallel receive lines are not possible to process in real-time using BS Capon in combination with our system configuration.

For Capon in element space, processing rates for high quality settings are still to high too support real-time cardiac ultrasound imaging. From Fig. \ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 100 ms (10 fps). Taking into account that we use complex data, but less samples per image, these results are similar to the performance reported by Chen et al. \cite{Chen2011}. For the 32 element array in Fig. (\ref{fig:benchUS32}) execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$).

A major part of the total execution time for element space Capon comes from solving the system of linear equations. This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Fig. \ref{fig:benchBS}, inverting the 3x3 beamspace covariance matrix is then barely visible. The time it takes to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time. If real-time is defined as 10 fps \cite{Chen2011} all execution times in both Fig. \ref{fig:bench} and \ref{fig:benchBS} are real-time. More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image. With this requirement the images formed using a 32, 64 and 96 element array have a real-time requirement of 125, 62.5 and 41.7 fps respectively using single line acquisition. Each benchmark plot has a line indicating this real-time requirement. With beamspace processing, all processing times for large subarrays are below this line.

@@

\section{Conclusion}\label{sec:con}
In this paper a parallel formulation and implementation of the element- and beam-space Capon beamformer have been presented, and
both beamformers have successfully been implemented on a GPU. For beamspace Capon beamforming processing rates supporting real-time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters. Image quality was investigated both in a simulated phantom and an \textit{in-vivo} cardiac dataset. These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects.

@@

\begin{IEEEbiography}{Jon Petter \AA{}sen}
Biography text here.
\end{IEEEbiography}

\begin{IEEEbiography}{Jo Inge Buskenes}
Biography text here.
\end{IEEEbiography}

\begin{IEEEbiography}{Carl-Inge Colombo Nilsen}
Biography text here.
\end{IEEEbiography}

@@

\begin{IEEEbiography}{Andreas Austeng}
Biography text here.
\end{IEEEbiography}

\begin{IEEEbiography}{Sverre Holm}
Biography text here.
\end{IEEEbiography}

>>>>

\documentclass[journal]{IEEEtran}

% Single column review mode:
%\documentclass[12pt,journal,onecolumn]{IEEEtran}
%\newcommand\CLASSINPUTbaselinestretch{2}
%\RequirePackage{calc}
%\RequirePackage{fp} 
% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
% Adding the 2cm gutter width to the odd/even side margins:
%\setlength\hoffset{0pt}
%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side

@@

\begin{abstract}
%\boldmath
Capon beamforming is associated with a high computational complexity, which limits its use as a real-time array processor in many applications. In this paper we present an implement of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting. To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the number of array elements increases. For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging is shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters. Image quality is investigated in videos of both a simulated phantom and an \textit{in-vivo} cardiac dataset. These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects, both in element- and beam-space.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

% denne må utelates i IUS-artikkelen
In recent work \cite{asen2012, Buskenes} we have applied a GPU-based Capon beamformer for both active sonar and for cardiac ultrasound imaging. Sonar has a real-time requirement less strict than medical ultrasound imaging, which means that most array configurations can be handled in real-time using the GPU. Medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates at a level where even a high-end GPU can not do the processing required for a straightforward implementation of the Capon beamformer. Since the execution time is a cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming have therefore recently been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we will make use of a beamspace transformation described by Nilsen et al. \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases. The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

@@

\begin{figure}
\centerline{\includegraphics[width=3.4in]{gfx/gpu_layout_vertical_lr.png}}
\caption{Calculation of Capon adaptive weights mapped to GPU architecture. A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups. Note that several samples can be processed per block of threads if $L$ is small. The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
\label{fig:gpulayout}
\end{figure}

@@

To reach this goal we start by showing how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
In time we have that
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculate from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ \cite{Buskenes}. However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$ (if $L=M/2$). Note that sliding will break independence along the sliding dimension, thus, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is increased to the maximum number of threads per block (currently 1024) with this approach. However, for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.

Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated. However, the GPU computes $\mat{\hat{R}}$ in such a way that the lower triangle can be found with little overhead \cite{Buskenes}. This also means that no special solver is needed in order to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if a symmetric solver could have saved some instructions.

@@

\begin{figure}
%\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
%\hfil
%\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_606.png}%
%\label{fig:benchUS64}}}
\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15$ cm deep sector. Execution time is plotted as a function of subarray length $L$. The plot shows execution times for the three steps listed in Section \ref{sec:meth}. Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
\label{fig:bench}
\end{figure}

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
\hfil
\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
\label{fig:benchBS96}}}
\caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$ cm deep sector. Execution time is plotted as a function of subarray length $L$.  If $L$ had been kept constant on $M/2$ and execution time was plotted against $N_b$, the plot would be similar to the one in Fig. \ref{fig:bench}. The variation in execution time is caused by adaptive kernel launch configurations.}
\label{fig:benchBS}
\end{figure*}
In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth). Problem sizes are calculated based on 64 and 96 element phased arrays with a 3.4 MHz center frequency. Assuming a system bandwidth of 80\% we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line. To correctly sample a $70\degree$ sector with 64 and 96 element arrays ($\lambda/2$ pitch), we need at least 80 and 120 receive beams respectively. The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in GPU memory. What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.

A benchmark of the implementation described in Section \ref{sec:meth} is presented in Fig. \ref{fig:bench}. The benchmark is based on a 64 elements phased array, using 80 receive beams. For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$. For $M=64$, $L=32$ and $K=2$ (Fig. \ref{fig:bench}), a typical setup for high quality cardiac imaging, processing throughput is 7 frames per second.

In Fig. \ref{fig:benchBS} we present similar benchmarks for the beamspace version of the Capon beamformer introduced in Section \ref{sec:bs}. Benchmarks are based on 64 and 96 elements phased arrays, using 80 and 120 transmit beams. The figure only present results with use of temporal averaging ($K=2$). For $M=64$, $L=32$, $K=2$ and $N_b=3$, processing throughput is 125 frames per second (8 ms). For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 77 frames per second (13 ms). In all four benchmarks there is a line indicating the time it takes to acquire the underlying image using single line acquisition. Hence, below this line we are capable of performing real-time processing when a single receive line is used per transmit line.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=3in]{gfx/simulation_slice.png} \label{fig:simulationSlice}}
\hfil
\subfloat[]{\includegraphics[width=3in]{gfx/invivo_slice.png} \label{fig:invivoSlice}}
}
\caption{Lateral slices taken from the images in Fig. \ref{fig:phantom} and Fig. \ref{fig:invivo}. a) From Fig. \ref{fig:phantom} at 93 mm constant depth, through the two bright point scatterers. b) From Fig. \ref{fig:invivo} at 96 mm range, through the tip of the mitral valve.}
\label{fig:slices}
\end{figure*}

To investigate the image quality of both the Capon and BS Capon beamformer a phantom was simulated using Field II \cite{Jensen1992} (Fig. \ref{fig:phantom}). The simulation was performed using a $\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth and 10 mm high phased array. The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm. The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB initially located at 90 mm depth. Speckle level is at 0 dB. The resulting image has been processed both using DAS with uniform weights, Capon and BS Capon. The presented image is the first frame in a series of images where motion is imposed on both circles, point targets and the probe. The entire simulation is presented in an attached video.

@@

To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Fig. \ref{fig:phantom} and \ref{fig:invivo} are included in Fig. \ref{fig:slices}. For the phantom, the cut goes through the two point targets at 93 mm constant depth, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant depth.

@@

\section{Trading resolution for speed}\label{sec:trade}
As seen from Fig. \ref{fig:bench}, execution time is reduced when $L$ is reduced. In Fig. \ref{fig:benchBS} we see that execution time is also reduced when moving to beamspace. So to gain an  increase in processing throughput we can reduce $L$ in element space or $N_b$ in beamspace. The question is then, how does this impact image quality? In Fig. \ref{fig:speed_res_trade}, the minimum distance between the two point targets in Fig. \ref{fig:phantom} (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with Capon) and $N_b$ (when processed with BS Capon). For the result in Fig. \ref{fig:speed_res_trade}, $d=0.01$ has been used on both methods.

For high $L$ and $N_b$ values the two methods are equal. However, for lower values, BS Capon provides higher resolution than what is possible with small subarrays. We see that the resolution for BS Capon gradually decreases due to an automatically increase in diagonal loading (See end of Section \ref{sec:bs}). If this is compensated for, BS Capon will provide resolution almost equal to Capon (with $L=M/2$) for all values except $N_b=1$ \cite{Nilsen2009}. For $N_b=1$ the resolution is worse than for DAS because of a triangular apodization caused by subarray averaging. In element space, resolution decays rapidly as a result of a reduced effective adaptive aperture. For $N_b=3$, BS Capon still provides approximately 1 mm improved resolution over DAS. However, note that resolution for the Capon beamformer highly depends on image SNR and the choice of parameters. Thus, different SNR scenarios and choices of parameters will give different results. It is also worth noting that Capon beamforming in this example performs worse than DAS for small subarrays. For small values of $L$, there are not enough degrees of freedom in order perfectly steer a zero at each point target.

@@

\section{Discussion}\label{sec:dis}
In this paper we have presented a parallel implementation of the Capon beamformer in both element- and beam-space for high resolution cardiac ultrasound imaging. The results show that it is now possible to apply BS Capon in real-time in cardiac ultrasound imaging by utilizing the power of GPUs. According to Fig. \ref{fig:benchBS} it is also possible to process two parallel receive lines per transmit (for both $M=64$ and $M=96$. $L=M/2$). However, these extra lines might be needed for increased sampling instead of increased frame rate when using the Capon beamformers. If parallel receive lines are used to obtain denser sampling, acquisition time remains the same but processing times are increased with a factor equal to the amount of oversampling. If parallel receive lines are used for increased frame rate, the acquisition time is lowered, and the processing time remains constant. Hence, four parallel receive lines are not possible to process in real-time using BS Capon (with $N_b=3$) in combination with our system configuration.

For Capon in element space, processing rates for high quality settings are still to high too support real-time cardiac ultrasound imaging. From Fig. \ref{fig:bench} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 140 ms (7 fps). Taking into account that we use complex data, but less samples per image, these results are similar to the performance reported by Chen et al. \cite{Chen2011}.

A major part of the total execution time for element space Capon comes from solving the system of linear equations. This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Fig. \ref{fig:benchBS}, inverting the 3x3 beamspace covariance matrix is then barely visible. The time it takes to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time. If real-time is defined as 10 fps \cite{Chen2011} almost all execution times in both Fig. \ref{fig:bench} and \ref{fig:benchBS} are real-time. More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image. With this requirement the images formed using a 64 and 96 element array have a real-time requirement of 62.5 and 41.7 fps respectively using single line acquisition. Each benchmark plot has a line indicating this real-time requirement. With beamspace processing, all processing times for large subarrays are below this line.

@@

\section{Conclusion}\label{sec:con}
In this paper a parallel formulation and implementation of the element- and beam-space Capon beamformer have been presented, and
both beamformers have successfully been implemented on a GPU. For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters. Image quality was investigated both in a simulated phantom and an \textit{in-vivo} cardiac dataset. These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects, both in element- and beam-space.

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jp.jpg}}]{Jon Petter \AA{}sen}
(S'12) was born in Porsgrunn, Norway in 1986. He received the M.S degree in computer science from the University of Oslo, Norway, in 2010. He is currently pursuing his Ph.D degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway. His research interests include adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics processing units (GPUs). 
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/ji.jpg}}]{Jo Inge Buskenes}
Biography text here.
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/ci.png}}]{Carl-Inge Colombo Nilsen}
(S’06–M’10) received the M.sc. and Ph.d. degrees in computer science from the University of Oslo, Norway, in 2005 and 2010. He is currently working at the University of Oslo as a postdoctoral research fellow. His research interests include signal and array processing for ultrasound imaging and other acoustic applications.
\end{IEEEbiography}

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/aa.jpg}}]{Andreas Austeng}
was born in Oslo, Norway, in 1970. He received the M.S. degree in physics in 1996 and the Ph.D. degree in computer science in 2001, both from the University of Oslo. since 2001, he has been working at the department of Informatics, University of Oslo, first as a postdoctoral research fellow and currently as an associate Professor. His research interests include signal processing for acoustical imaging.
\end{IEEEbiography}

\newpage

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sh.jpg}}]{Sverre Holm}
(M’82-SM’02) was born in Oslo, Norway, in 1954. He received the M.S. and Ph.D. degrees in electrical engineering from the Norwegian Institute of Technology, Trondheim, in 1978 and 1982, respectively. His academic experience includes Yarmouk University in Jordan (1984-1986) and the Norwegian Institute of Technology and later the University of Oslo (1989-1994) as an adjunct professor. Since 1995, he has been a professor of signal processing in the department of Informatics at the University of Oslo. In 2002, he was elected a member of the Norwegian Academy of Technological Sciences, and in 2009, he became an adjunct professor in the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab, Trondheim. His industry experience includes GE Vingmed Ultrasound, Horten, Norway (1990-1994). He worked with Sonitor Technologies (2000-2005), where he developed ultrasonic technology that is the core of the indoor positioning system described in \textit{Scientific Americanin} 2008 as "a positioning system that goes where GPS can’t." He has published about 130 publications internationally, plus about 10 patents. He has had sabbaticals at GE Global Research, NY (1998), and the Waves and 
Acoustics Laboratory of ESPCI, Paris (2008-2009). He was an associate editor of the \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control} from 1997 to 2002. His research interests are medical ultrasound imaging, modeling of tissue-ultrasound interaction, and ultrasound positioning.
\end{IEEEbiography}
